---
title: "Early-Pandemic COVID-19 Epidemiologic Analysis"
format: html
---

# Study Data

This study uses a 10,000-record sample of reported COVID-19 cases restricted to the early pre-vaccine study period from March 15 through December 10, 2020.

```{r}
# Load COVID-19 surveillance sample
covid <- read.csv(
    "Data/covid_early_pandemic_sample_10k.csv",
    stringsAsFactors = FALSE
)

dim(covid)
names(covid)
head(covid)
```

# Analytic Cohort

The analysis is restricted to cases with known exposure and outcome status. Exposure is defined by `medcond_yn`, indicating whether a pre-existing medical condition was reported. The outcome is defined by `hosp_yn`, indicating a hospitalization was reported.

`medcond_yn` and `hosp_yn` are coded as {`Yes`, `No`, `Unknown`, `Missing`}. Only `Yes` and `No` records for either the exposure or outcome are kept for the complete-case analysis.

```{r}
analysis <- covid[
    covid$medcond_yn %in% c("Yes", "No") &
    covid$hosp_yn %in% c("Yes", "No"),
]

dim(analysis)
```

The resulting complete-case analytic cohort contains 1,464 cases.

# Crude Association

## Exposure-Outcome Table

We first estimate the crude association between the exposure (reported pre-existing medical condition) and outcome (hospitalization).

```{r}
exposure_outcome <- table(
    analysis$medcond_yn,
    analysis$hosp_yn
)

exposure_outcome
```

The table's orientation lists exposure as rows and outcome as columns.

## Risk Ratio and Risk Difference

Hospitalization risk is estimated separately among exposed and
unexposed cases.

```{r}
# Extract cells from exposure-outcome table
a <- exposure_outcome["Yes", "Yes"]
b <- exposure_outcome["Yes", "No"]
c <- exposure_outcome["No", "Yes"]
d <- exposure_outcome["No", "No"]

# Hospitalization risks
risk_exposed <- a / (a + b)
risk_unexposed <- c / (c + d)

# Effect measures
risk_ratio <- risk_exposed / risk_unexposed
risk_difference <- risk_exposed - risk_unexposed

risk_exposed
risk_unexposed
risk_ratio
risk_difference
```

Among cases in the complete-case analytic cohort, hospitalization risk was approximately 35.3% among those with a reported pre-existing medical condition and 3.6% among those without one. The crude risk ratio was 9.86, corresponding to an absolute risk difference of approximately 31.7 percentage points.

## 95% Confidence Interval for Risk Ratio

The conventional Wald confidence interval for the risk ratio is
defined as:

$$
SE[\ln(RR)] =
\sqrt{
\frac{1}{a}
-\frac{1}{a+b}
+\frac{1}{c}
-\frac{1}{c+d}
}
$$

```{r}
se_log_rr <- sqrt(
    (1 / a) - (1 / (a + b)) +
    (1 / c) - (1 / (c + d))
)

log_rr <- log(risk_ratio)

log_rr_lower <- log_rr - (1.96 * se_log_rr)
log_rr_upper <- log_rr + (1.96 * se_log_rr)

rr_lower <- exp(log_rr_lower)
rr_upper <- exp(log_rr_upper)

c(
    RR = risk_ratio,
    Lower_95_CI = rr_lower,
    Upper_95_CI = rr_upper
)
```

The crude risk ratio was 9.86 (95% CI: 6.72--14.47).

## 95% Confidence Interval for Risk Difference

For two independent proportions the standard error is defined as:

$$
SE(RD) =
\sqrt{
\frac{R_1(1-R_1)}{n_1}
+
\frac{R_0(1-R_0)}{n_0}
}
$$

```{r}
n_exposed <- a + b
n_unexposed <- c + d

se_rd <- sqrt(
    (risk_exposed * (1 - risk_exposed) / n_exposed) +
    (risk_unexposed * (1 - risk_unexposed) / n_unexposed)
)

rd_lower <- risk_difference - (1.96 * se_rd)
rd_upper <- risk_difference + (1.96 * se_rd)

c(
    RD = risk_difference,
    Lower_95_CI = rd_lower,
    Upper_95_CI = rd_upper
)
```

The crude risk difference was 31.7 percentage points (95% CI: 27.9--35.4 percentage points).

These estimates are crude and unadjusted. These confidence intervals treat observations as independent.

# Descriptive Analysis

Age and sex were hypothesized to be potential confounders based on the DAG and literature review. Consequently, their distributions are described according to both exposure and outcome status before adjusted analysis.

## Age

### Age by Exposure

```{r}
age_exposure <- table(
    analysis$age_group,
    analysis$medcond_yn
)

age_exposure

round(
    prop.table(age_exposure, margin = 2) * 100,
    1
)
```

The column percentages describe the age distribution within each exposure group, $P(Age\mid E)$.

### Hospitalization by Age

```{r}
age_outcome <- table(
    analysis$age_group,
    analysis$hosp_yn
)

age_outcome

round(
    prop.table(age_outcome, margin = 1) * 100,
    1
)
```

The row percentages describe hospitalization status within each age group, $P(Y\mid Age)$.

## Sex

### Sex by Exposure

```{r}
sex_exposure <- table(
    analysis$sex,
    analysis$medcond_yn
)

sex_exposure

round(
    prop.table(sex_exposure, margin = 2) * 100,
    1
)
```

The column percentages describe the sex distribution within each exposure group, $P(Sex\mid E)$.

### Hospitalization by Sex

```{r}
sex_outcome <- table(
    analysis$sex,
    analysis$hosp_yn
)

sex_outcome

round(
    prop.table(sex_outcome, margin = 1) * 100,
    1
)
```

The row percentages describe hospitalization status within each sex group, $P(Y\mid Sex)$.

# Adjusted Analysis

The adjusted analysis needs to account for the age and sex variables because they were identified as potential confounders. Two records in the complete-case analytic cohort have sex coded as `Missing` or `Unknown` and have been excluded from the analysis.

```{r}
adjusted_analysis <- analysis[
    analysis$sex %in% c("Female", "Male"),
]

dim(adjusted_analysis)
table(adjusted_analysis$sex)
table(adjusted_analysis$age_group)

# Encode variables for regression
adjusted_analysis$hospitalized <- factor(
    adjusted_analysis$hosp_yn,
    levels = c("No", "Yes")
)

adjusted_analysis$medcond <- factor(
    adjusted_analysis$medcond_yn,
    levels = c("No", "Yes")
)

adjusted_analysis$sex_factor <- factor(
    adjusted_analysis$sex,
    levels = c("Female", "Male")
)

adjusted_analysis$age_factor <- factor(
    adjusted_analysis$age_group,
    levels = c(
        "0 - 9 Years",
        "10 - 19 Years",
        "20 - 29 Years",
        "30 - 39 Years",
        "40 - 49 Years",
        "50 - 59 Years",
        "60 - 69 Years",
        "70 - 79 Years",
        "80+ Years"
    )
)

levels(adjusted_analysis$hospitalized)
levels(adjusted_analysis$medcond)
levels(adjusted_analysis$sex_factor)
levels(adjusted_analysis$age_factor)

table(adjusted_analysis$hospitalized)
table(adjusted_analysis$medcond)
table(adjusted_analysis$sex_factor)
table(adjusted_analysis$age_factor)
```

The resulting adjusted-analysis cohort contained 1,462 cases. Hospitalization was modeled with No as the reference category and Yes as the event. Reported pre-existing medical condition used No as the reference category, sex used Female as the reference category, and age group was modeled categorically with 0--9 Years as the reference category.

## Logistic Regression

A multivariable logistic regression model was fit with hospitalization as the outcome, reported pre-existing medical condition as the exposure, and age group and sex as prespecified adjustment variables.

The logistic model estimates conditional log odds of hospitalization. Exponentiated coefficients therefore represent adjusted odds ratios rather than adjusted risk ratios.

```{r}
logistic_model <- glm(
    hospitalized ~ medcond + age_factor + sex_factor,
    data = adjusted_analysis,
    family = binomial(link = "logit")
)

summary(logistic_model)
```

### Logistic regression limitations

Our outcome is binary, so logistic regression is an obvious modeling choice. However, the logit link models the log odds of the outcome rather than the risk directoy. The fitted model can be written as:

$$
\operatorname{logit}(p_i) = \beta_0 + \beta_E E_i + \sum_{k=1}^{8}\beta_{A_k}A_{ik} + \beta_S S_i,
$$

where

$$
p_i = P(Y_i = 1 \mid E_i, A_i, S_i).
$$

Here, $E_i$ represents reported pre-existing medical condition stuatus, $A_{ik}$ represents the eight indicator variables from the non-reference age groups, and $S_i$ represents sex.

The exposure coefficient $\beta_E$ is estimated on the log-odds scale. Exponentiating this coefficient gives the adjusted odds ratio:

$$
OR_{\text{adjusted}} = e^{\beta_E},
$$

rather than an adjusted risk ratio, which was selected as teh primary measure for this analysis. Consequently, the adjusted risk ratio will be estimated from the fitted logistic regression model using regression standardization.