---
title: "Early-Pandemic COVID-19 Epidemiologic Analysis"
format: html
---

# Study Data

This study uses 10,000 samples of reported COVID-19 cases restricted to the early pre-vaccine study period from March 15 through December 10, 2020.

```{r}
# Load COVID-19 surveillance sample
covid <- read.csv(
    "Data/covid_early_pandemic_sample_10k.csv",
    stringsAsFactors = FALSE
)

dim(covid)
names(covid)
head(covid)
```

# Analytic Cohort

The analysis is restricted to cases with known exposure and outcome status. Exposure is defined by `medcond_yn`, indicating whether a pre-existing medical condition was reported. The outcome is defined by `hosp_yn`, indicating a hospitalization.

The record label set is {`Yes`, `No`, `Unknown`, `Missing`}. Only `Yes` and `No` records for either the exposure or outcome are kept for the complete-case analysis.

```{r}
analysis <- covid[
    covid$medcond_yn %in% c("Yes", "No") &
    covid$hosp_yn %in% c("Yes", "No"),
]

dim(analysis)
```

The resulting complete-case analytic cohort contains 1,464 cases.

# Crude Association

## Exposure-Outcome Table

We first estimate the crude association between the exposure (reported pre-existing medical condition) and outcome (hospitalization).

```{r}
exposure_outcome <- table(
    analysis$medcond_yn,
    analysis$hosp_yn
)

exposure_outcome
```

The table's orientation lists exposure as rows and outcome as columns.

## Risk Ratio and Risk Difference

Hospitalization risk is estimated separately among exposed and
unexposed cases.

```{r}
# Extract cells from exposure-outcome table
a <- exposure_outcome["Yes", "Yes"]
b <- exposure_outcome["Yes", "No"]
c <- exposure_outcome["No", "Yes"]
d <- exposure_outcome["No", "No"]

# Hospitalization risks
risk_exposed <- a / (a + b)
risk_unexposed <- c / (c + d)

# Effect measures
risk_ratio <- risk_exposed / risk_unexposed
risk_difference <- risk_exposed - risk_unexposed

risk_exposed
risk_unexposed
risk_ratio
risk_difference
```

Among cases in the complete-case analytic cohort, hospitalization risk was approximately 35.3% among those with a reported pre-existing medical condition and 3.6% among those without one. The crude risk ratio was 9.86, corresponding to an absolute risk difference of approximately 31.7 percentage points.

## 95% Confidence Interval for Risk Ratio

The conventional Wald confidence interval for the risk ratio is
defined as:

$$
SE[\ln(RR)] =
\sqrt{
\frac{1}{a}
-\frac{1}{a+b}
+\frac{1}{c}
-\frac{1}{c+d}
}
$$

```{r}
se_log_rr <- sqrt(
    (1 / a) - (1 / (a + b)) +
    (1 / c) - (1 / (c + d))
)

log_rr <- log(risk_ratio)

log_rr_lower <- log_rr - (1.96 * se_log_rr)
log_rr_upper <- log_rr + (1.96 * se_log_rr)

rr_lower <- exp(log_rr_lower)
rr_upper <- exp(log_rr_upper)

c(
    RR = risk_ratio,
    Lower_95_CI = rr_lower,
    Upper_95_CI = rr_upper
)
```

The crude risk ratio was 9.86 (95% CI: 6.72-14.47).

## 95% Confidence Interval for Risk Difference

For two independent proportions the standard error is defined as:

$$
SE(RD) =
\sqrt{
\frac{R_1(1-R_1)}{n_1}
+
\frac{R_0(1-R_0)}{n_0}
}
$$

```{r}
n_exposed <- a + b
n_unexposed <- c + d

se_rd <- sqrt(
    (risk_exposed * (1 - risk_exposed) / n_exposed) +
    (risk_unexposed * (1 - risk_unexposed) / n_unexposed)
)

rd_lower <- risk_difference - (1.96 * se_rd)
rd_upper <- risk_difference + (1.96 * se_rd)

c(
    RD = risk_difference,
    Lower_95_CI = rd_lower,
    Upper_95_CI = rd_upper
)
```

The crude risk difference was 31.7 percentage points (95% CI: 27.9-35.4 percentage points).

These estimates are crude and unadjusted. These confidence intervals treat observations as independent.

# Descriptive Analysis

Age and sex were hypothesized to be potential confounders based on the DAG and literature review. Consequently, their distributions are described according to both exposure and outcome status before adjusted analysis.

## Age

### Age by Exposure

```{r}
age_exposure <- table(
    analysis$age_group,
    analysis$medcond_yn
)

age_exposure

round(
    prop.table(age_exposure, margin = 2) * 100,
    1
)
```

The column percentages describe the age distribution within each exposure group, $P(Age\mid E)$.

### Hospitalization by Age

```{r}
age_outcome <- table(
    analysis$age_group,
    analysis$hosp_yn
)

age_outcome

round(
    prop.table(age_outcome, margin = 1) * 100,
    1
)
```

The row percentages describe hospitalization status within each age group, $P(Y\mid Age)$.

## Sex

### Sex by Exposure

```{r}
sex_exposure <- table(
    analysis$sex,
    analysis$medcond_yn
)

sex_exposure

round(
    prop.table(sex_exposure, margin = 2) * 100,
    1
)
```

The column percentages describe the sex distribution within each exposure group, $P(Sex\mid E)$.

### Hospitalization by Sex

```{r}
sex_outcome <- table(
    analysis$sex,
    analysis$hosp_yn
)

sex_outcome

round(
    prop.table(sex_outcome, margin = 1) * 100,
    1
)
```

The row percentages describe hospitalization status within each sex group, $P(Y\mid Sex)$.